---
title: "PM566 Final Project"
author: "Karisa Ke"
output: 
    html_document:
        toc: TRUE
        toc_float: TRUE
---

This is my PM566 Final Project website.

```{r setup, message=FALSE, echo=FALSE, warning=FALSE}
library(data.table)
library(tidyverse)
library(dplyr)
library(plotly)
library(DT)
library(knitr)
library(ggplot2)
library(reshape2)

library(scales) 
```




```{r, message=FALSE, echo=FALSE, warning=FALSE}
housing <- read_csv("housing.csv")

haversine <- function(lat1, lon1, lat2, lon2) {
  R <- 6371  # Radius of Earth in km
  phi1 <- lat1 * (pi / 180)
  phi2 <- lat2 * (pi / 180)
  delta_phi <- (lat2 - lat1) * (pi / 180)
  delta_lambda <- (lon2 - lon1) * (pi / 180)
  
  a <- sin(delta_phi / 2)^2 + cos(phi1) * cos(phi2) * sin(delta_lambda / 2)^2
  c <- 2 * atan2(sqrt(a), sqrt(1 - a))
  d <- R * c
  return(d * 1000)  # Convert km to m
}

coordinates <- list(
  'Los Angeles' = c(34.0522, -118.2437),
  'San Diego' = c(32.7157, -117.1611),
  'San Jose' = c(37.3382, -121.8863),
  'San Francisco' = c(37.7749, -122.4194)
)


for (city in names(coordinates)) {
  lat <- coordinates[[city]][1]
  lon <- coordinates[[city]][2]
  housing[[paste('Distance to', city)]] <- mapply(haversine,housing$latitude, housing$longitude, lat, lon)
}

```



```{r, message=FALSE, echo=FALSE, warning=FALSE}
#  this is for the first  question
numeric_housing <- housing[sapply(housing, is.numeric)]

cor_matrix <- cor(numeric_housing, use = "complete.obs")

melted_cor_matrix <- melt(cor_matrix)


p <- ggplot(melted_cor_matrix, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1, 1), space = "Lab", 
                       name="Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.title.x = element_blank(),
        axis.title.y = element_blank())


```



#  this is for the second question


summary(housing)


income_40th_percentile <- quantile(housing$median_income, 0.40)
income_60th_percentile <- quantile(housing$median_income, 0.60)


filtered_data <- housing %>%
  filter(median_income >= income_40th_percentile & median_income <= income_60th_percentile)

p <- ggplot(filtered_data, aes(x = median_income, y = median_house_value, color = median_house_value)) +
  geom_point(alpha = 0.5) +
  scale_color_gradient(low = "blue", high = "red", 
                       breaks = pretty_breaks(n = 5)(range(filtered_data$median_house_value)), 
                       labels = scales::label_comma()) + 
  geom_smooth(method = 'gam', formula = y ~ s(x, bs = "cs"), color = "black") +
  scale_x_continuous(labels = scales::label_comma()) +
  scale_y_continuous(labels = scales::label_comma()) + # Add this line to format y axis
  theme_minimal()




p
ggplotly(p)





# this is for third question

fig <- plot_ly(data = housing, x = ~longitude, y = ~latitude, z = ~median_house_value, 
               type = 'scatter3d', mode = 'markers',
               marker = list(size = 2, 
                             color = ~median_house_value, 
                             colorscale = 'Bluered',  
                             colorbar = list(title = 'House Price'),
                             opacity = 0.8))

fig <- fig %>% layout(title = '3D Spatial Distribution of House Prices',
                      scene = list(xaxis = list(title = 'Longitude'),
                                   yaxis = list(title = 'Latitude'),
                                   zaxis = list(title = 'Price')))


fig




 
# this is for fourth question

housing$`Distance to Los Angeles` <- housing$`Distance to Los Angeles` * 0.000621371

housing$`Distance to San Diego` <- housing$`Distance to San Diego` * 0.000621371

housing$`Distance to San Francisco` <- housing$`Distance to San Francisco` * 0.000621371

housing$`Distance to San Jose` <- housing$`Distance to San Jose` * 0.000621371


la_near_housing <- housing %>% 
  filter(`Distance to Los Angeles` <= 200)


p <- ggplot(la_near_housing, aes(x = `Distance to Los Angeles`, y = median_house_value, color = median_house_value)) +
  geom_point(alpha = 0.5) +
  scale_color_gradient(low = "blue", high = "red",
                       breaks = pretty_breaks(n = 5)(range(la_near_housing$median_house_value)), 
                       labels = label_comma()) +
  geom_smooth(method = 'gam', formula = y ~ s(x, bs = "cs"), color = "black") +
  scale_x_continuous(labels = label_comma()) +
  theme_minimal()


ggplotly(p)




# for sd 
sd_near_housing <- housing %>% 
  filter(`Distance to San Diego` <= 150)


p_sd <- ggplot(sd_near_housing, aes(x = `Distance to San Diego`, y = median_house_value, color = median_house_value)) +
  geom_point(alpha = 0.5) +
  scale_color_gradient(low = "blue", high = "red",
                       breaks = pretty_breaks(n = 5)(range(sd_near_housing$median_house_value)), 
                       labels = label_comma()) +
  geom_smooth(method = 'gam', formula = y ~ s(x, bs = "cs"), color = "black") +
  scale_x_continuous(labels = label_comma()) +
  theme_minimal()


ggplotly(p_sd)

# this is for fifth question




fig <- plot_ly(data = housing, x = ~longitude, y = ~latitude, z = ~housing_median_age, 
               type = 'scatter3d', mode = 'markers',
               marker = list(size = 2, 
                             color = ~housing_median_age, 
                             colorscale = 'Bluered',  
                             colorbar = list(title = 'House age'),
                             opacity = 0.8))

fig <- fig %>% layout(title = '3D Spatial Distribution of housing_median_age',
                      scene = list(xaxis = list(title = 'Longitude'),
                                   yaxis = list(title = 'Latitude'),
                                   zaxis = list(title = 'housing_median_age')))


fig



summary(housing)

p <- ggplot(housing, aes(x = housing_median_age)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
  labs(title = "Histogram of Housing Median Age",
       x = "Housing Median Age",
       y = "Frequency") +
  theme_minimal()

# Print the plot
p
```

