---
title: "PM566 Final Project"
author: "Karisa Ke"
output: 
    html_document:
        toc: TRUE
        toc_float: TRUE
---

# Introduction
The Californian housing market stands as a multifaceted landscape shaped by a myriad of socio-economic and geographical factors. Understanding the intricate dynamics governing property values and development across this region is pivotal, not only for real estate professionals but also for policymakers and individuals seeking insights into housing trends and patterns.

This project embarks on an in-depth exploration into the Californian housing market using a robust dataset sourced from the 1990 census. The primary objective is to unravel the intricate relationships between various factors influencing housing prices, with a particular focus on income levels, geographical coordinates, and housing age distribution.



# Interactive visualizations
```{r setup, message=FALSE, echo=FALSE, warning=FALSE}
library(data.table)
library(tidyverse)
library(dplyr)
library(plotly)
library(DT)
library(knitr)
library(ggplot2)
library(reshape2)

library(scales) 
```


```{r, message=FALSE, echo=FALSE, warning=FALSE}
housing <- read_csv("data_folder/housing.csv")

haversine <- function(lat1, lon1, lat2, lon2) {
  R <- 6371  # Radius of Earth in km
  phi1 <- lat1 * (pi / 180)
  phi2 <- lat2 * (pi / 180)
  delta_phi <- (lat2 - lat1) * (pi / 180)
  delta_lambda <- (lon2 - lon1) * (pi / 180)
  
  a <- sin(delta_phi / 2)^2 + cos(phi1) * cos(phi2) * sin(delta_lambda / 2)^2
  c <- 2 * atan2(sqrt(a), sqrt(1 - a))
  d <- R * c
  return(d * 1000)  # Convert km to m
}

coordinates <- list(
  'Los Angeles' = c(34.0522, -118.2437),
  'San Diego' = c(32.7157, -117.1611),
  'San Jose' = c(37.3382, -121.8863),
  'San Francisco' = c(37.7749, -122.4194)
)


for (city in names(coordinates)) {
  lat <- coordinates[[city]][1]
  lon <- coordinates[[city]][2]
  housing[[paste('Distance to', city)]] <- mapply(haversine,housing$latitude, housing$longitude, lat, lon)
}

```

### 1. Correlation Between Variables
To assess the strength of correlation between the continuous variables within the dataset, an initial analysis was conducted to validate the expected relationships between specific variables. This correlation matrix presents an overall examination of the interrelationships between variables. 

```{r, message=FALSE, echo=FALSE, warning=FALSE}
#  this is for the first  question
numeric_housing <- housing[sapply(housing, is.numeric)]

cor_matrix <- cor(numeric_housing, use = "complete.obs")

melted_cor_matrix <- melt(cor_matrix)


p <- ggplot(melted_cor_matrix, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1, 1), space = "Lab", 
                       name="Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.title.x = element_blank(),
        axis.title.y = element_blank())

p
```

### 2. House Value for the Medians of Median_Income
A focused analysis was undertaken to validate whether individuals falling within the 40%-60% percentile range of median_income also exhibit house values within the corresponding median range of median_house_value data. The 40%-60% percentile of median_income corresponds to \$30,000 to \$40,000 USD. From the trend line, incomes ranging between \$30,000 and \$40,000 have house values spanning from \$180,000 to \$210,000, which is close to the calculated 40%-60% percentile range of median_house_value.

```{r, message=FALSE, echo=FALSE, warning=FALSE}
#  this is for the second question


income_40th_percentile <- quantile(housing$median_income, 0.40)
income_60th_percentile <- quantile(housing$median_income, 0.60)


filtered_data <- housing %>%
  filter(median_income >= income_40th_percentile & median_income <= income_60th_percentile)

p <- ggplot(filtered_data, aes(x = median_income, y = median_house_value, color = median_house_value)) +
  geom_point(alpha = 0.5) +
  scale_color_gradient(low = "blue", high = "red", 
                       breaks = pretty_breaks(n = 5)(range(filtered_data$median_house_value)), 
                       labels = scales::label_comma()) + 
  geom_smooth(method = 'gam', formula = y ~ s(x, bs = "cs"), color = "black") +
  scale_x_continuous(labels = scales::label_comma()) +
  scale_y_continuous(labels = scales::label_comma()) + # Add this line to format y axis
  theme_minimal()

p

```


### 3. Location House Value
This 3D projection using longitude, latitude, and median_house_value offers a perspective on how location influences housing prices. The dynamic nature of the interactive model further allows users to explore and navigate through the geographical landscape, gaining an understanding of how specific locations correspond to varying median_house_values.

```{r, message=FALSE, echo=FALSE, warning=FALSE}
# this is for third question

fig <- plot_ly(data = housing, x = ~longitude, y = ~latitude, z = ~median_house_value, 
               type = 'scatter3d', mode = 'markers',
               marker = list(size = 2, 
                             color = ~median_house_value, 
                             colorscale = 'Bluered',  
                             colorbar = list(title = 'House Price'),
                             opacity = 0.8))

fig <- fig %>% layout(title = '3D Spatial Distribution of House Prices',
                      scene = list(xaxis = list(title = 'Longitude'),
                                   yaxis = list(title = 'Latitude'),
                                   zaxis = list(title = 'Price')))

fig

```


### 4. House Values Around Los Angeles Area
This scatter plot coupled with a trend line illustrates the fluctuation in house values concerning the distance from the center of Los Angeles

```{r, message=FALSE, echo=FALSE, warning=FALSE}
# this is for fourth question

housing$`Distance to Los Angeles` <- housing$`Distance to Los Angeles` * 0.000621371

housing$`Distance to San Diego` <- housing$`Distance to San Diego` * 0.000621371

housing$`Distance to San Francisco` <- housing$`Distance to San Francisco` * 0.000621371

housing$`Distance to San Jose` <- housing$`Distance to San Jose` * 0.000621371


la_near_housing <- housing %>% 
  filter(`Distance to Los Angeles` <= 200)


p <- ggplot(la_near_housing, aes(x = `Distance to Los Angeles`, y = median_house_value, color = median_house_value)) +
  geom_point(alpha = 0.5) +
  scale_color_gradient(low = "blue", high = "red",
                       breaks = pretty_breaks(n = 5)(range(la_near_housing$median_house_value)), 
                       labels = label_comma()) +
  geom_smooth(method = 'gam', formula = y ~ s(x, bs = "cs"), color = "black") +
  scale_x_continuous(labels = label_comma()) +
  theme_minimal()

p
```


```{r, message=FALSE, echo=FALSE, warning=FALSE}
# for sd 
sd_near_housing <- housing %>% 
  filter(`Distance to San Diego` <= 150)


p_sd <- ggplot(sd_near_housing, aes(x = `Distance to San Diego`, y = median_house_value, color = median_house_value)) +
  geom_point(alpha = 0.5) +
  scale_color_gradient(low = "blue", high = "red",
                       breaks = pretty_breaks(n = 5)(range(sd_near_housing$median_house_value)), 
                       labels = label_comma()) +
  geom_smooth(method = 'gam', formula = y ~ s(x, bs = "cs"), color = "black") +
  scale_x_continuous(labels = label_comma()) +
  theme_minimal()


```


### 5. Pattern of Housing Median Age Across California
Even though the correlation between housing_median_age and geographical location (longitude/latitude) according to the correlation matrix, the 3D projection below offers additional insights: Across geographical coordinates, there are both new and old houses on each coordinate, indicating a evenly spread real estate development pattern

```{r, message=FALSE, echo=FALSE, warning=FALSE}
# this is for fifth question

fig <- plot_ly(data = housing, x = ~longitude, y = ~latitude, z = ~housing_median_age, 
               type = 'scatter3d', mode = 'markers',
               marker = list(size = 2, 
                             color = ~housing_median_age, 
                             colorscale = 'Bluered',  
                             colorbar = list(title = 'House age'),
                             opacity = 0.8))

fig <- fig %>% layout(title = '3D Spatial Distribution of housing_median_age',
                      scene = list(xaxis = list(title = 'Longitude'),
                                   yaxis = list(title = 'Latitude'),
                                   zaxis = list(title = 'housing_median_age')))


fig

p <- ggplot(housing, aes(x = housing_median_age)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
  labs(title = "Histogram of Housing Median Age",
       x = "Housing Median Age",
       y = "Frequency") +
  theme_minimal()
```



# Download the Report
The writtern report is available to be downloaded here: 
[Download PDF](./PM566_Final_Project_ written_report.pdf)